---
title: "Water Quality Exploratory Data Analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE }
library(tidyverse)
library(readxl)
library(ggplot2)
library(ggridges)
library(corrplot)
library(car)
library(psych)
library(writexl)
```

### Set Working Directory
```{r, echo=FALSE }

working_dir <- "C:/BF510/final_project/"

```

## Read in Water Quality Dataset

At this point, the selected water quality dataset has been filtered to only include sampling dates with complete specific conductivity, temperature, total dissolved solids, DO, and TDS data with outliers removed.

filtered_water_sites.csv generated from raw water quality data in water_data_preprocessing.rmd

```{r , echo=FALSE}
water_quality_data <- read.csv(paste0(working_dir, "/data/" ,"filtered_water_sites.csv"))

#Convert DOSAT and DO columns to numeric

water_quality_data$DOSAT <- as.numeric(water_quality_data$DOSAT)
water_quality_data$DO <- as.numeric(water_quality_data$DO)

head(water_quality_data)
```

## Create Exploratory Plots of the Water Quality Dataset

### Specific Conductance


View SPCOND data by year, clustered by watershed (boxplot)
```{r , echo=FALSE}

spcond_1 <- ggplot(water_quality_data, aes(x=Datayear, y=SPCOND, group = Datayear)) +
    geom_boxplot()+
    facet_wrap(~Watershed)+
    scale_x_continuous(breaks=c(2005,2010,2015,2020),labels=c("2005","2010","2015","2020"))+
    labs(x="Year", y=" Specific Conductance (µS/cm)", , title = "Specific Conductance by Year and Watershed -- Boxplots") +
    theme_minimal()
spcond_1
```

View SPCOND data by watershed (density plots)

```{r , echo=FALSE}

spcond_2 <- ggplot(water_quality_data, aes(x=SPCOND, y=Watershed)) +
    geom_density_ridges(rel_min_height = 0.01)+
    facet_wrap(~Datayear)+
    labs(x="Specific Conductance (µS/cm)", y="Watershed", title = "Specific Conductance by Watershed -- Density Plots") +
    theme_minimal()
spcond_2
```


### Temperature


View TEMP data by year, clustered by watershed (boxplot)

```{r , echo=FALSE}

temp_1 <- ggplot(water_quality_data, aes(x=Datayear, y=TEMP, group = Datayear)) +
    geom_boxplot()+
    facet_wrap(~Watershed)+
    scale_x_continuous(breaks=c(2005,2010,2015,2020),labels=c("2005","2010","2015","2020"))+
    labs(x="Year", y="Temperature (°C)", title="Temperature by Year and Watershed -- Boxplots") +
    theme_minimal()
temp_1
```

View TEMP data by watershed (density plots)

```{r , echo=FALSE}

temp_2 <- ggplot(water_quality_data, aes(x=TEMP, y=Watershed)) +
    geom_density_ridges(rel_min_height = 0.01)+
    facet_wrap(~Datayear)+
    labs(x="Temperature (°C)", y="Watershed", title = "Temperature by Watershed -- Density Plots") +
    theme_minimal()
temp_2
```


### Total Dissolved solids

Total dissolved solids (TDS), calculated from specific conductivity (mg/l)

View TDS data by year, clustered by watershed (boxplot)

```{r , echo=FALSE}

tds_1 <- ggplot(water_quality_data, aes(x=Datayear, y=TDS, group = Datayear)) +
    geom_boxplot()+
    facet_wrap(~Watershed)+
    scale_x_continuous(breaks=c(2005,2010,2015,2020),labels=c("2005","2010","2015","2020"))+
    labs(x="Year", y="TDS (mg/l)", title="Total Dissolved Solids by Year and Watershed -- Boxplots") +
    theme_minimal()
tds_1
```

View TDS data by watershed (density plots)

```{r , echo=FALSE}

tds_2 <- ggplot(water_quality_data, aes(x=TDS, y=Watershed)) +
    geom_density_ridges(rel_min_height = 0.01)+
    facet_wrap(~Datayear)+
    labs(x="TDS (mg/l)", y="Watershed", title = "Total Dissolved Solids by Watershed -- Density Plots") +
    theme_minimal()
tds_2
```


### Dissolved Oxygen

View DO data by year, clustered by watershed (boxplot)

```{r , echo=FALSE}

do_1 <- ggplot(water_quality_data, aes(x=Datayear, y=DO, group = Datayear)) +
    geom_boxplot()+
    facet_wrap(~Watershed)+
    scale_x_continuous(breaks=c(2005,2010,2015,2020),labels=c("2005","2010","2015","2020"))+
    labs(x="Year", y="Dissolved Oxygen (mg/l)", title="Dissolved Oxygen by Year and Watershed -- Boxplots") +
    theme_minimal()

do_1
```

View DO data by watershed (density plots)

```{r , echo=FALSE}

do_2 <- ggplot(water_quality_data, aes(x=DO, y=Watershed)) +
    geom_density_ridges(rel_min_height = 0.01)+
    facet_wrap(~Datayear)+
    labs(x="Dissolved Oxygen (mg/l)", y="Watershed", title = "Dissolved Oxygen by Watershed -- Density Plots") +
    theme_minimal()
do_2
```


### Dissolved Oxygen Saturation

View DOSAT data by year, clustered by watershed (boxplot)

```{r , echo=FALSE}

dosat_1 <- ggplot(water_quality_data, aes(x=Datayear, y=DOSAT, group = Datayear)) +
    geom_boxplot()+
    facet_wrap(~Watershed)+
    scale_x_continuous(breaks=c(2005,2010,2015,2020),labels=c("2005","2010","2015","2020"))+
    labs(x="Year", y="Dissolved Oxygen Saturation (%)", title="Dissolved Oxygen Saturation by Year and Watershed -- Boxplots") +
    scale_y_continuous(limits=c(0,200),breaks=seq(0,200,25)) +
    theme_minimal()

dosat_1
```

View DOSAT data by watershed (density plots)

```{r , echo=FALSE}

dosat_2 <- ggplot(water_quality_data, aes(x=DOSAT, y=Watershed)) +
geom_density_ridges(rel_min_height = 0.01)+
facet_wrap(~Datayear)+
labs(x="Dissolved Oxygen Saturation (%)", y="Watershed", title = "Dissolved Oxygen Saturation by Watershed -- Density Plots") +
theme_minimal()

dosat_2
```


### pH


View PH data by year, clustered by watershed (boxplot)

```{r , echo=FALSE}

ph_1 <- ggplot(water_quality_data, aes(x=Datayear, y=PH, group = Datayear)) +
    geom_boxplot()+
    facet_wrap(~Watershed)+
    scale_x_continuous(breaks=c(2005,2010,2015,2020),labels=c("2005","2010","2015","2020"))+
    labs(x="Year", y="pH", title="pH by Year and Watershed -- Boxplots") +
    scale_y_continuous(limits=c(0,200),breaks=seq(0,200,25)) +
    theme_minimal()

ph_1
```

View PH data by watershed (density plots)

```{r , echo=FALSE}

ph_2 <- ggplot(water_quality_data, aes(x=PH, y=Watershed)) +
geom_density_ridges(rel_min_height = 0.01)+
facet_wrap(~Datayear)+
labs(x="pH", y="Watershed", title = "pH by Watershed -- Density Plots") +
theme_minimal()

ph_2
```

#### Save plots as png files 

```{r , echo=FALSE}
plots <- list("spcond_1"=spcond_1,"spcond_2"=spcond_2,"do_1"=do_1,"do_2"=do_2,"dosat_1"=dosat_1,"dosat_2"=dosat_2,"temp_1"=temp_1,"temp_2"=temp_2,"ph_1"=ph_1,"ph_2"=ph_2)
variables <- names(plots)

for (i in variables) {
  ggsave(filename = paste0(working_dir, "results/exploratory_plots/", i, ".png"), 
         plot = plots[[i]], width = 12, height = 16, dpi = 300,
         bg="white")
}

```

## Multicolinearity Analysis

Use to see whether any water quality variables are significantly highly correlated

### Create Subsetted Data Frame with Variables of Interest only

```{r}

water_vars_subset_df <- water_quality_data %>% 
     select("nSPCOND", "nTDS", "nPH", "nTEMP", "nDO", "nDOSAT", "nDEPTH", "Datayear")

```

### Create Pearson Correlation Matrix and Plot Result
```{r}
cor_matrix <- cor(water_vars_subset_df)

corrplot(cor_matrix, method = "circle")
```
- DO and DOSAT are highly multicolinear, which is expected since one is calculated from the other.
- TDS and SPCOND are highly multicolinear, which is expected since one is calculated from the other.

### Calculate Variance Inflation Factors
```{r}
lm_model <-lm(nTEMP ~ Datayear + nDEPTH + nSPCOND + nTDS + nPH + nDO + nDOSAT, data = water_vars_subset_df)

vif_values <- vif(lm_model)

print(vif_values)
```

There are very high VIFs for SPCOND, TDS, DO, and DOSAT. This is expected since TDS is calculated from SPCOND, and DOSAT is based on DO. I will remove DOSAT and TDS and re-calculate the VIFs. 


```{r}
lm_model2 <-lm(nTEMP ~ Datayear + nDEPTH + nSPCOND + nPH + nDO , data = water_vars_subset_df)

vif_values2 <- vif(lm_model2)

print(vif_values2)
```

The VIFs are ~ 1 for all remaining variables, suggesting little to no multicolinearity. I will retain these in the statistical models.

## Summary Statistics for Selected Water Quality and EJ Metrics

### Create Table with Summary Statistics Water Quality Metrics included in EJ Zone Proximity Analysis

Included statistics for log10-transformed depth and SPCOND since they are used in the GLMMs.

```{r}

# Create log10-transformed depth and SPCOND columns
water_vars_subset_df$logDEPTH <- log10(water_vars_subset_df$nDEPTH)
water_vars_subset_df$logSPCOND <- log10(water_vars_subset_df$nSPCOND)

# Remove DOSAT and TDS

water_vars_subset_df_2 <- select(water_vars_subset_df, -c("nTDS", "nDOSAT"))

water_qual_summary_table <- water_vars_subset_df_2 %>% 
    select(-c("Datayear")) %>%
    describe(.) %>%
    select(-c("vars")) %>%
    mutate(across(where(is.numeric), ~round(., 3)))

# Remove the "n" from row names
rownames(water_qual_summary_table) <- gsub("^n", "", rownames(water_qual_summary_table))


# Create new column to store row names, add to beginning of data frame

water_qual_summary_table <- water_qual_summary_table %>%
  mutate(variables = rownames(water_qual_summary_table)) %>%
  select(variables, everything())

# Write table to csv file
write_xlsx(water_qual_summary_table, paste0(working_dir, "results/summary_statistics/EJ_zone_proximity_variables_summary_stats.xlsx"))

```

### Read in Data Frame with Water Quality and EJ Demographics data combined

Data set generated using filtered water quality data set from above, MassDEP EJ geospatial and demographics data set, and ArcGISPro Model.  
```{r}
ej_water_qual_data <- read.csv(paste0(working_dir, "/data/" ,"water_quality_EJ_data_merged.csv"))
```

### Generate Summary Table for just EJ demographics data

Includes selected EJ variables used in analyses.
```{r}

# Variables to select
ej_zone_variables <- c("BG_MHHI", "LIMENGHHPCT", "PCT_MINORITY", "TOTAL_POP")

# Select information from distinct EJ neighborhoods and only select columns with demographics information
ej_demog_info <- ej_water_qual_data %>%
                        distinct(GEOGRAPHICAREANAME, .keep_all = TRUE) %>%
                        select(., ej_zone_variables)



ej_demog_summary_table <- ej_demog_info %>% 
    describe(.) %>%
    select(-c("vars")) %>%
    mutate(across(where(is.numeric), ~round(., 3))) 

# Round median household income row to 2 decimals

ej_demog_summary_table["BG_MHHI", ] <- round(ej_demog_summary_table["BG_MHHI", ], 2)


# Create new column to store row names, add to beginning of data frame

ej_demog_summary_table <- ej_demog_summary_table %>%
  mutate(variables = rownames(ej_demog_summary_table)) %>%
  select(variables, everything())

# Write table to csv file

write_xlsx(ej_demog_summary_table, paste0(working_dir, "results/summary_statistics/Within_EJ_demographics_variables_summary_stats.xlsx"))


```


### Generate Summary Table for Water Quality Metrics used in Within EJ analysis

Includes selected EJ variables used in analyses.
```{r}

# Create log10-transformed depth and SPCOND columns
ej_water_qual_data$logDEPTH <- log10(ej_water_qual_data$nDEPTH)
ej_water_qual_data$logSPCOND <- log10(ej_water_qual_data$nSPCOND)

# List of variables to summarize and subset data frame to only include those

ej_zone_variables2 <- c("logSPCOND","nSPCOND","nPH","nTEMP", "nDO", "nDEPTH", "logDEPTH")


ej_water_qual_data_2 <- ej_water_qual_data %>% select(.,ej_zone_variables2)
                        
ej_water_qual_summary_table <- ej_water_qual_data_2 %>% 
    describe(.) %>%
    select(-c("vars")) %>%
    mutate(across(where(is.numeric), ~round(., 3))) 


# Remove the "n" from row names
rownames(ej_water_qual_summary_table) <- gsub("^n", "", rownames(ej_water_qual_summary_table))

# Create new column to store row names, add to beginning of data frame

ej_water_qual_summary_table <- ej_water_qual_summary_table %>%
  mutate(variables = rownames(ej_water_qual_summary_table)) %>%
  select(variables, everything())

# Write table to csv file

write_xlsx(ej_water_qual_summary_table, paste0(working_dir, "results/summary_statistics/Within_EJ_water_quality_variables_summary_stats.xlsx"))


```


### Calculate Number of Unique Water Monitoring Sites in Each Zone Across MA

Data set generated with ArcGISPro geospatial data processing pipeline. 
```{r}
## Read in Data set with Zone labels 

water_qual_zone_buffers <- read.csv(paste0(working_dir, "/data/", "water_quality_zone_labels2.csv"))

sites_by_zone_count <- water_qual_zone_buffers %>%
    group_by(.,Zone_Label) %>%
    summarize(Site_Count = n_distinct(UNIQUE_ID)) %>%
    as.data.frame()

```

