---
title: "Water Quality Statistical Models"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(tidyverse)
library(glmmTMB)
library(purrr)
library(broom.mixed)
library(DHARMa)
library(emmeans)
library(writexl)
```


### Set Working Directory
```{r, echo=FALSE }

working_dir <- "C:/BF510/final_project/"

```


## Read in processed water quality datasets 

```{r echo=FALSE}

water_qual_zone_buffers <- read.csv(paste0(working_dir, "/data/", "water_quality_zone_labels2.csv"))

water_qual_ej_zone_only <- read.csv(paste0(working_dir, "/data/", "water_quality_EJ_data_merged.csv"))

```

## Data setup

```{r echo=FALSE}

# Make EJ zone proximity labels a factor with the "Greater_Than_3_Miles" category as reference
water_qual_zone_buffers$Zone_Label <- factor(water_qual_zone_buffers$Zone_Label,  
                          levels = c("Greater_Than_3_Miles", "Inside_EJ", "Within_1_Mile", "Within_3_Miles"),
                          labels = c("Greater Than 3 Miles", "Inside EJ", "Within 1 Mile", "Within 3 Miles"))

water_qual_ej_zone_only$Zone_Label <- factor(water_qual_ej_zone_only$Zone_Label,  
                          levels = c("Inside_EJ"),
                          labels = c("Inside EJ"))

# Set municipality column to be a factor
water_qual_ej_zone_only$MUNICIPALITY <- factor(water_qual_ej_zone_only$MUNICIPALITY)

# Set sampling site to be a factor
water_qual_zone_buffers$UNIQUE_ID <- factor(water_qual_zone_buffers$UNIQUE_ID)

water_qual_ej_zone_only$UNIQUE_ID <- factor(water_qual_ej_zone_only$UNIQUE_ID)

# Set year column to be a factor
water_qual_zone_buffers$Datayear <- factor(water_qual_zone_buffers$Datayear)
water_qual_ej_zone_only$Datayear <- factor(water_qual_ej_zone_only$Datayear)


# Create log 10-transformed Depth column to include in models as a covariate
water_qual_zone_buffers$logDEPTH <- log10(water_qual_zone_buffers$nDEPTH)
water_qual_ej_zone_only$logDEPTH <- log10(water_qual_ej_zone_only$DEPTH)

# Create log 10-transformed SPCOND to normalize values
water_qual_zone_buffers$logSPCOND <- log10(water_qual_zone_buffers$nSPCOND)
water_qual_ej_zone_only$logSPCOND <- log10(water_qual_ej_zone_only$nSPCOND)


```


## EJ Zone Proximity Analysis

The goal of this analysis is to see if proximity to EJ neighborhoods as classified by the four categories (Greater Than 3 Miles, Inside EJ, Within 1 Mile, Within 3 Miles) is significantly associated with water quality measures. 

I am using UNIQUE_ID to account for random effects, which represents the water sampling stations. I also included data year and log 10-transformed water depth as covariates in the models.

For the distribution families, I used gamma distributions with a log link function for all water quality variables except temperature since they can handle skewed data well (mentioned in https://doi.org/10.2166/wqrj.2022.028).
For temperature, I used a gaussian/normal distribution since the data has only a small left skew and AIC/BIC values were slightly better for skew-normal and tweedie distributions 


### Function to apply models to water quality variables
```{r}

fit_water_quality_models_1 <- function(variables, data) {
  
 # List to store model results
  model_summaries <- list()
  
  for (variable in variables) {
    # Define model formula
    formula <- as.formula(paste(variable, "~ Zone_Label + Datayear + logDEPTH + (1|UNIQUE_ID)"))

    # Fit model
    model <- glmmTMB(formula, data = data, family = Gamma(link = "log"))
    
    # Get the full summary of the model
    full_summary <- summary(model)
    
    # Tidy model output and create 95% CI column
    tidy_summary <- broom.mixed::tidy(model, conf.int = TRUE, conf.level = 0.95) %>%
      mutate(
        CI_95 = paste0(round(conf.low, 3), " - ", round(conf.high, 3)),
        BH_FDR = p.adjust(p.value, method = "BH", n = length(p.value))      
      ) %>%
      select(term, estimate, std.error, statistic, p.value, BH_FDR, CI_95) %>%
      as.data.frame()
    
    # Store both summaries in the list for the current variable
    model_summaries[[variable]] <- list(
      model_objects = model,
      full_summary = full_summary,
      tidy_summary = tidy_summary
    )
  }
  
  return(model_summaries)  # Return the list of both summaries
}

```


```{r echo=FALSE}

water_vars <- c("logSPCOND", "nDO", "nPH")

model_outputs_1 <- lapply(water_vars, fit_water_quality_models_1, data = water_qual_zone_buffers)


```


## TEMP 
```{r echo=FALSE}

temp_zone_model <- glmmTMB(nTEMP ~ Zone_Label + Datayear + logDEPTH + (1|UNIQUE_ID), data = water_qual_zone_buffers, family = gaussian(link="identity")) 
        
summary(temp_zone_model)
```

### Process temperature results
```{r}

# Tidy  model output and create 95% CI column
temp_zone_tidy_summary <- broom.mixed::tidy(temp_zone_model, conf.int = TRUE, conf.level = 0.95) %>%
    mutate(
    CI_95 = paste0(round(conf.low, 3), " - ", round(conf.high, 3)),
        BH_FDR = p.adjust(p.value, method = "BH", n = length(p.value))  
    ) %>%
    select(term, estimate, std.error, statistic, p.value, BH_FDR, CI_95) %>%
    as.data.frame()

```

### Write results to xlsx file
```{r}

out <- list("Temp" = temp_zone_tidy_summary, "logSPCOND" = model_outputs_1[[1]]$logSPCOND$tidy_summary, 
            "DO" = model_outputs_1[[2]]$nDO$tidy_summary, "PH" = model_outputs_1[[3]]$nPH$tidy_summary)
write_xlsx(out, paste0(working_dir, "results/model_outputs/EJ_zone_proximity_analysis/EJ_Zone_Proximity_Analysis_Results.xlsx"))

```

## Generate Predictive and Diagnostic Plots of the EJ Zone Proximity Results

### Assessing Model Fit

Here, I will use QQ plots to detect overall deviations from the expected data distributions.

I set the number of simulations to n = 1000 to stabilize the simulated values. 

### SPCOND

```{r}
diagnostics_spcond_1 <- simulateResiduals(fittedModel = model_outputs_1[[1]]$logSPCOND$model_objects, plot = FALSE, n = 1000)

png(filename = paste0(working_dir, "results/model_outputs/EJ_zone_proximity_analysis/model_output_plots/spcond_model_output_residual_plot.png"), width = 12, height = 12, res = 300, units = "in") 
spcond_resid_plot1 <- plot(diagnostics_spcond_1)
spcond_resid_plot1
dev.off()
```

### DO

```{r}
diagnostics_DO_1 <- simulateResiduals(fittedModel = model_outputs_1[[2]]$nDO$model_objects, plot = FALSE, n = 1000)

png(filename = paste0(working_dir, "results/model_outputs/EJ_zone_proximity_analysis/model_output_plots/do_model_output_residual_plot.png"), width = 12, height = 12, res = 300, units = "in") 
do_resid_plot1 <- plot(diagnostics_DO_1)
do_resid_plot1
dev.off()
```

### PH

```{r}
diagnostics_ph_1 <- simulateResiduals(fittedModel = model_outputs_1[[3]]$nPH$model_objects, plot = FALSE, n = 1000)

png(filename = paste0(working_dir, "results/model_outputs/EJ_zone_proximity_analysis/model_output_plots/ph_model_output_residual_plot.png"), width = 12, height = 12, res = 300, units = "in") 
ph_resid_plot1 <- plot(diagnostics_ph_1)
ph_resid_plot1
dev.off()
```

### TEMP
```{r}

diagnostics_temp_1 <- simulateResiduals(fittedModel = temp_zone_model, plot = FALSE, n = 1000)

png(filename = paste0(working_dir, "results/model_outputs/EJ_zone_proximity_analysis/model_output_plots/temp_model_output_residual_plot.png"), width = 12, height = 12, res = 300, units = "in") 
temp_resid_plot1 <- plot(diagnostics_temp_1)
temp_resid_plot1
dev.off()
```

### Estimating Means of Water Quality variables Based on Model Fit using emmeans()

Focusing on logSPCOND and pH since there were significant associations in those analyses. 

Also includes pairwise comparisons between mean logSPCOND values between categories, as calculated by a Tukey test. 

### logSPCOND
```{r}

# Calculate estimated marginal means of logSPCOND for each Zone category along with pairwise comparisons between zones

logspcond_means <- emmeans(model_outputs_1[[1]]$logSPCOND$model_objects, specs = pairwise ~ Zone_Label, type = "response", plotIt=FALSE)

# Plot estimated marginal means across the four categories

logspcond_emmeans_plot <- plot(logspcond_means,
                            horizontal=FALSE, 
                            xlab = "Mean logSPCOND",
                            ylab = "EJ Neighborhood Proximity",
                            colors = c("orange")) +
                            ggtitle("Mean logSPCOND By EJ Neighborhood Proximity") +
                            theme_classic()+
                            theme(plot.title = element_text(hjust=0.5)) 

logspcond_emmeans_plot

# Create data frame from pairwise comparison test and add in sample sizes
spcond_emmeans_df <- as.data.frame(logspcond_means$emmeans) %>% mutate(df = 8335)
spcond_emmeans_tukey_df <- as.data.frame(logspcond_means$contrasts) %>% mutate(df = 8335)

```


### PH
```{r}

# Calculate estimated marginal means of pH for each Zone category along with pairwise comparisons between zones

ph_means <- emmeans(model_outputs_1[[3]]$nPH$model_objects, specs = pairwise ~ Zone_Label, type = "response", plotIt=FALSE)

# Plot estimated marginal means across the four categories

ph_emmeans_plot <- plot(ph_means,
                            horizontal=FALSE, 
                            xlab = "Mean pH",
                            ylab = "EJ Neighborhood Proximity") +
                            ggtitle("Mean pH By EJ Neighborhood Proximity") +
                            theme_classic() +
                            theme(plot.title = element_text(hjust=0.5))
                            
ph_emmeans_plot

# Create data frames from emmeans calculations pairwise comparison test and add in sample sizes
ph_emmeans_df <- as.data.frame(ph_means$emmeans) %>% mutate(df = 8335)
ph_emmeans_tukey_df <- as.data.frame(ph_means$contrasts) %>% mutate(df = 8335)

```

### Save all posthoc plots and results 
```{r}

#Save all data tables to an excel spreadsheet
zone_posthoc_out <- list("logSPCOND_emmeans" = spcond_emmeans_df, "logSPCOND_posthoc_tukey_test" = spcond_emmeans_tukey_df,
"pH_emmeans" = ph_emmeans_df, "pH_posthoc_tukey_test" = ph_emmeans_tukey_df)
write_xlsx(zone_posthoc_out, paste0(working_dir, "results/model_outputs/EJ_zone_proximity_analysis/EJ_Zone_Proximity_Posthoc_Analysis_Results.xlsx"))


#Save as pdf

#Create list of posthoc analysis plots (was unable to save residual plots with this method)

posthoc_plots_1 <- list("logSPCOND Estimated Marginal Means EJ Zone Proximity Model Plot" = logspcond_emmeans_plot,
                        "PH Estimated Marginal Means EJ Zone Proximity Model Plot" = ph_emmeans_plot)




pdf(paste0(working_dir, "results/model_outputs/EJ_zone_proximity_analysis/model_output_plots/EJ_Zone_Proximity_Posthoc_Analysis_Results_Plots.pdf"), width = 8, height = 6)
for (plot_name in names(posthoc_plots_1)) {
  print(posthoc_plots_1[[plot_name]])  # Print each histogram to the PDF
}
dev.off()
```




```


## Within EJ Zone Analysis

The goal of this analysis is to see if different EJ metrics are associated with water quality measures within EJ neighborhood blocks.

I accounted for random effects across municipality and the sampling site with a nested structure. 

### Function to apply models to water quality variables

```{r echo=FALSE}

fit_water_quality_models_2 <- function(variables, data) {
  
 # List to store model results
  model_summaries <- list()
  
  for (variable in variables) {
    # Define model formula
    formula <- as.formula(paste(variable, "~ BG_MHHI + LIMENGHHPCT + PCT_MINORITY + TOTAL_POP + logDEPTH + Datayear +(1 | MUNICIPALITY/UNIQUE_ID)"))

    # Fit model
    model <- glmmTMB(formula, data = data, family = Gamma(link = "log"))
    
    # Get the full summary of the model
    full_summary <- summary(model)
    
    # Tidy model output and create 95% CI column
    tidy_summary <- broom.mixed::tidy(model, conf.int = TRUE, conf.level = 0.95) %>%
      mutate(
        CI_95 = paste0(round(conf.low, 3), " - ", round(conf.high, 3)),
        BH_FDR = p.adjust(p.value, method = "BH", n = length(p.value)) 
      ) %>%
      select(term, estimate, std.error, statistic, p.value, BH_FDR, CI_95) %>%
      as.data.frame()
    
    # Store both summaries in the list for the current variable
    model_summaries[[variable]] <- list(
      model_objects = model,
      full_summary = full_summary,
      tidy_summary = tidy_summary
    )
  }
  
  return(model_summaries)  # Return the list of both summaries
}
```

## Run models for all water quality variables except temperature 

Used a slightly different model for temperature

```{r echo=FALSE}

model_outputs_2 <- lapply(water_vars, fit_water_quality_models_2, data = water_qual_ej_zone_only)

```

## Temperature Model

```{r echo=FALSE}

temp_ej_model <- glmmTMB(nTEMP ~ BG_MHHI + LIMENGHHPCT + PCT_MINORITY + TOTAL_POP + logDEPTH + Datayear + (1 | MUNICIPALITY/UNIQUE_ID), data = water_qual_ej_zone_only, family = gaussian(link="identity")) 

summary(temp_ej_model)

```

### Process temperature results
```{r}

# Tidy  model output and create 95% CI column
temp_ej_tidy_summary <- broom.mixed::tidy(temp_ej_model, conf.int = TRUE, conf.level = 0.95) %>%
    mutate(
    CI_95 = paste0(round(conf.low, 3), " - ", round(conf.high, 3)),
        BH_FDR = p.adjust(p.value, method = "BH", n = length(p.value)) 
    ) %>%
    select(term, estimate, std.error, statistic, p.value, BH_FDR, CI_95) %>%
    as.data.frame()
    
```

```{r}
out2 <- list("Temp" = temp_ej_tidy_summary, "logSPCOND" = model_outputs_2[[1]]$logSPCOND$tidy_summary,  
            "DO" = model_outputs_2[[2]]$nDO$tidy_summary, "PH" = model_outputs_2[[3]]$nPH$tidy_summary)
write_xlsx(out2, paste0(working_dir, "results/model_outputs/Within_EJ_zone_analysis/Within_EJ_Analysis_Results.xlsx"))

```

## Generate Predictive and Diagnostic Plots of the Within EJ Zone Results

### Assessing Model Fit

Here, I will use QQ plots to detect overall deviations from the expected data distributions.

I set the number of simulations to n = 1000 to stabilize the simulated values. 

### SPCOND

```{r}
diagnostics_spcond_2 <- simulateResiduals(fittedModel = model_outputs_2[[1]]$logSPCOND$model_objects, plot = FALSE, n = 1000)

png(filename = paste0(working_dir, "results/model_outputs/Within_EJ_zone_analysis/model_output_plots/spcond_model_output_residual_plot.png"), width = 12, height = 12, res = 300, units = "in") 
plot(diagnostics_spcond_2)
dev.off()
```

### DO

```{r}
diagnostics_DO_2 <- simulateResiduals(fittedModel = model_outputs_2[[2]]$nDO$model_objects, plot = FALSE, n = 1000)

png(filename = paste0(working_dir, "results/model_outputs/Within_EJ_zone_analysis/model_output_plots/do_model_output_residual_plot.png"), width = 12, height = 12, res = 300, units = "in") 
plot(diagnostics_DO_2)
dev.off()
```

### PH

```{r}
diagnostics_ph_2 <- simulateResiduals(fittedModel = model_outputs_2[[3]]$nPH$model_objects, plot = FALSE, n = 1000)

png(filename = paste0(working_dir, "results/model_outputs/Within_EJ_zone_analysis/model_output_plots/ph_model_output_residual_plot.png"), width = 12, height = 12, res = 300, units = "in") 
plot(diagnostics_ph_2)
dev.off()
```

### TEMP
```{r}

diagnostics_temp_2 <- simulateResiduals(fittedModel = temp_ej_model, plot = FALSE, n = 1000)

png(filename = paste0(working_dir, "results/model_outputs/Within_EJ_zone_analysis/model_output_plots/temp_model_output_residual_plot.png"), width = 12, height = 12, res = 300, units = "in") 
plot(diagnostics_temp_2)
dev.off()
```